Predicting Conflict Escalation

### 2023-04-25 docs/hes_final/article.qmd
get tables to work in two column article

---
title: "Predicting Conflict Escalation"
author:
  - name: Diana Kolusheva
    email: email@h
    affiliation: Harvard University
  - name: Chad Stoughton
    email: email@h
    affiliation: Harvard University
  - name: Evan Wheeler
    email: email@h
    affiliation: Harvard University
lang: en
format: pdf
classoption: twocolumn
bibliography: bibliography.bibtex
embed-resources: true
title-block-banner: true
abstract: |
  In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to COVID-19 response efforts globally. Innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children. Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need. Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944. Today, much of the work in this area has continued to focus on country or region-specific models to predict conflict.

@@

Prior to training, the GDELT dataset was split into one 512xn matrix for each country and month where n is the number of articles in the dataset published in that month related to that country. Each of these matrices was then split into several random subsamples of 512x50. Samples for countries in which n<50 were padded, and multiple samples were taken for countries in which n >50. This resampling dramatically increased the number of training samples by acting as though the same month had occurred multiple times while still maintaining uniqueness of each sample. Simultaneously, limiting the size of each sample to 50 articles reduced the dimensionality of each sample, reducing the risk of overfitting.

The encoder was then trained using these samples, with the time-shifted spike metric as its target. Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons, and its weights were frozen. The encoder was then used to scan each of the original 512xn matrices, resulting in a length-16 vector for each 512x50 subsample. This reduced the dimensions of each sample to 16x(n/50).

@@

TODO: Table (for numerical metrics) and plots (for roc_auc) for 2 models

>>>>

---
title: "Predicting Conflict Escalation"
author:
  - name: Diana Kolusheva
    email: email@h
    affiliation: Harvard University
  - name: Chad Stoughton
    email: email@h
    affiliation: Harvard University
  - name: Evan Wheeler
    email: email@h
    affiliation: Harvard University
lang: en
format:
  pdf:
    classoption:
    - twocolumn
bibliography: bibliography.bibtex
include-in-header:
  - text: |
      \usepackage{float}
      \makeatletter
      \let\oldlt\longtable
      \let\endoldlt\endlongtable
      \def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
      \def\longtable@i[#1]{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt[#1]
      }
      \def\longtable@ii{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt
      }
      \def\endlongtable{\endoldlt
      \end{minipage}
      \twocolumn
      \end{figure}}
      \makeatother
embed-resources: true
title-block-banner: true
abstract: |
  In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to COVID-19 response efforts globally. Innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children. Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need. Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944. Today, much of the work in this area has continued to focus on country or region-specific models to predict conflict.

@@

Prior to training, the GDELT dataset was split into one $512 x n$ matrix for each country and month where n is the number of articles in the dataset published in that month related to that country. Each of these matrices was then split into several random subsamples of $512 x 50$. Samples for countries in which $n < 50$ were padded, and multiple samples were taken for countries in which $n > 50$. This resampling dramatically increased the number of training samples by acting as though the same month had occurred multiple times while still maintaining uniqueness of each sample. Simultaneously, limiting the size of each sample to 50 articles reduced the dimensionality of each sample, reducing the risk of overfitting.

The encoder was then trained using these samples, with the time-shifted spike metric as its target. Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons, and its weights were frozen. The encoder was then used to scan each of the original $512 x n$ matrices, resulting in a length-16 vector for each $512 x 50$ subsample. This reduced the dimensions of each sample to $16 x (n/50)$.

@@

TODO: Table (for numerical metrics) and plots (for roc_auc) for 2 models
```{python}
#| echo: false
#| label: tbl-planets
#| tbl-cap: Planets

from IPython.display import Markdown
from tabulate import tabulate
table = [["Sun",696000,1989100000],
         ["Earth",6371,5973.6],
         ["Moon",1737,73.5],
         ["Mars",3390,641.85]]
Markdown(tabulate(
  table,
  headers=["Planet","R (km)", "mass (x 10^29 kg)"]
))
```
